Explore tumor cell classification with CellAssign

### 2024-04-24 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md

# Responding to code reviews

After you file your PR, the Data Lab team will assign a reviewer to your pull request.
We aim to begin the process of code review within three business days of you filing the PR.
You can expect that each PR will undergo several rounds of code review and response to review

@@

On the top of the PR page, you will see four tabs:

<figure markdown="span">
    ![Pull request tabs.](../../img/respond-to-review-tabs.png){width="800"}
</figure>

<div class="grid" markdown>

- **`Conversation`**
    - This tab shows the main page of the PR.
    Here, you can find the original description of the PR along with any comments left on the PR, both reviewer comments and any comments you leave in response.
    - You can also find any [resolved comments](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/commenting-on-a-pull-request#resolving-conversations) here, if you need to refer to them again.
- **`Commits`**
    - This tab shows a list of all commits added to the branch that is being reviewed.
    Your reviewer may use this page to explore changes in individual commits.

@@

<!-- comment to force two grid columns -->

- **`Checks`**
    - This tab shows results from certain types of code checks that have been automatically set up by the Data Lab; you can ignore this tab.
- **`Files changed`**
    - This tab shows a line-by-line overview of content you have added and removed in your branch.
    Changes shown here should reflect the difference between files in your branch and the base branch your PR is targeting. 
This should be the `main` branch of the `AlexsLemonade/OpenScPCA-analysis` repository. 
    - Referring to this tab is a great way to quickly see all your changes, as well as respond to [reviewer suggestions or comments](#inline-comments).

### 2024-04-24 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md

# Responding to code reviews

Code review is essentially a conversation between you and your reviewer.
To ensure that this process goes smoothly and with fewer opportunities for miscommunication, we strongly recommend reading through all of your reviewer's comments before diving into review:

@@

</div>

## Types of reviewer comments

When reviewing your pull request, there are several types of comments your reviewer may leave for you to respond to.
<!-- Please refer to [our documentation on example reviews](STUB_LINK example reviews). -->

!!! tip "Reading review comments"
    Unresolved file-level and inline review comments will appear both in the `Conversation` and in the `Files Changed` tabs.

@@

You can commit the batched suggestions at once with the `Commit suggestions` button that will appear at the top of the `Files changed` tab:

<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-commitbatch.png){width="600"}
</figure>

Read this GitHub documentation for more about [applying suggested changes while responding to review](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/incorporating-feedback-in-your-pull-request).

## Responding to review comments

When making code changes in response to review, you need to make sure your local repository is staying in-sync with any changes made directly on GitHub.
Follow these steps when responding to review to avoid [merge conflicts](../creating-pull-requests/resolve-merge-conflicts.md):

1. Before making any code changes, _pull_ your branch from GitHub.
This will ensure you are up-to-date with any code changes made on GitHub, including committed review suggestions or updates from merging the upstream `main` branch into your pull request branch.
    <figure markdown="span">
        ![Pull in GitKraken.](../../img/respond-to-review-pull.png){width="600"}
    </figure>

1. Make changes, commit, and push as you normally would.
Any changes you push to the branch with an open PR will automatically be reflected in the PR.

### 2024-04-24 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md

- In the first comment, the reviewer suggested that the code author add a new line to load the `readr` library.<br><br><br><br>
- In the second comment, the reviewer did not formally suggest a code change, but left a comment recommending to use `file.path()`.
(Learn more about [structuring your scripts](../analysis-modules/script-structure.md)!)<br><br>
- In the third comment, the reviewer suggested that the code author remove a line that loads the `dplyr` library.

<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-inline.png){width="600"}
</figure>

</div>

## Replying to comments

You can directly reply to review comments on GitHub.
For example, you might ask your reviewer to clarify a particular comment, or ask for more guidance about how to implement the review.
Or, you might just want to leave your reviewer a helpful note of some kind!

The way to responding to an existing comment differs depending on which PR tab you are leaving the comment from:

- If you are in the **`Files changed`** tab, use the button `Add single comment` to reply (_do not use `Start a review`_):
    <figure markdown="span">
        ![Reply to a review comment from the `Files changed` tab.](../../img/respond-to-review-comment-filed-changed.png){width="600"}
    </figure>

- If you are in the **`Conversation`** tab, use the button `Comment`.
    <figure markdown="span">
        ![Reply to a review comment from the `Conversation` tab.](../../img/respond-to-review-comment-conversation.png){width="600"}
    </figure>

When leaving a comment, you can either click `Add single comment` or `Start a review`.
You should click `Add single comment`:

## Working with suggestion comments

@@

## Responding to review comments

When making code changes in response to review, you need to make sure your local repository is staying in-sync with any changes made directly on GitHub.
Follow these steps when responding to review to avoid [merge conflicts](../creating-pull-requests/resolve-merge-conflicts.md):

1. Before making any code changes, _pull_ your branch from GitHub.
This will ensure you are up-to-date with any code changes made on GitHub, including committed review suggestions or updates from merging the upstream `main` branch into your pull request branch.
    <figure markdown="span">
        ![Pull in GitKraken.](../../img/respond-to-review-pull.png){width="600"}
    </figure>

1. Make changes, commit, and push as you normally would.
Any changes you push to the branch with an open PR will automatically be reflected in the PR.
    - When committing changes in response to review, the smaller the better!
    Review will go more smoothly if each commit addresses only one review comment.

@@

## Re-requesting review

Once you feel that you have addressed your reviewer's comments, you can [re-request review by clicking the cycle icon next to their handle on the right side of the PR page](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/incorporating-feedback-in-your-pull-request#re-requesting-a-review).

<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-x.png){width="400"}
</figure>

You can also post a comment in your PR summarizing your response to review, which will help your reviewer navigate your code changes:

- Which review comments did you address?
    - Is there more information you can tell your reviewer that will help them understand your code changes?
- Which review comments, if any, did you not address?
    - Was it because the review comment itself was confusing, or because you weren't sure how to address it?
    Communicating this with your reviewer will help you both out!

### 2024-04-24 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md

For example, this file-level review comment requests that the code author comment and document a script:

<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-filecomment.png){width="600"}
</figure>

### Inline comments

Your reviewer can also leave inline comments with feedback about specific lines of code:

@@

The way to responding to an existing comment differs depending on which PR tab you are leaving the comment from:

- If you are in the **`Files changed`** tab, use the button `Add single comment` to reply (_do not use `Start a review`_):
    <figure markdown="span">
        ![Reply to a review comment from the `Files changed` tab.](../../img/respond-to-review-comment-filed-changed.png){width="600"}
    </figure>

- If you are in the **`Conversation`** tab, use the button `Comment`.
    <figure markdown="span">
        ![Reply to a review comment from the `Conversation` tab.](../../img/respond-to-review-comment-conversation.png){width="600"}
    </figure>

(Learn more about [structuring your scripts](../analysis-modules/script-structure.md)!)

### Suggestion comments

These comments can also include _suggestions_ for code changes, that you can incorporate into your PR directly via on GitHub.

For example, below you are two suggestions review comments on an R script:

- In the first comment, the reviewer suggested that the code author add a new line to load the `readr` library.
- In the second comment, the reviewer suggested that the code author remove a line that loads the `dplyr` library.

<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-suggestions.png){width="600"}
</figure>

Suggestion comments have two associated buttons boxed in the image above: `Commit suggestion` and `Add suggestion to batch`.

### 2024-04-25 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md

# Responding to code reviews

Code review is essentially a conversation between you and your reviewer about your proposed changes. 
This document includes some example review comments and guidance on how to respond to review comments and suggestions. 
To ensure that this process goes smoothly and with fewer opportunities for miscommunication, we strongly recommend reading through all of your reviewer's comments before diving into review:

@@

- If you are in the **`Files changed`** tab, use the button `Add single comment` to reply (_do not use `Start a review`_):
    <figure markdown="span">
        ![Reply to a review comment from the `Files changed` tab.](../../img/respond-to-review-comment-filed-changed.png){width="600"}
    </figure>

- If you are in the **`Conversation`** tab, use the button `Comment`.
    <figure markdown="span">
        ![Reply to a review comment from the `Conversation` tab.](../../img/respond-to-review-comment-conversation.png){width="600"}
    </figure>

### Suggestion comments

Reviewer comments can also include _suggestions_ for code changes, that you can incorporate into your PR directly on GitHub.

For example, below are two suggestions left as comments on an R script:

### 2024-04-25 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md

# Responding to code reviews

Code review is essentially a conversation between you and your reviewer about your proposed changes.
This document includes some example review comments and guidance on how to respond to review comments and suggestions.

To ensure that code review goes smoothly and with fewer opportunities for miscommunication, we strongly recommend reading through all of your reviewer's comments before diving into review:

@@

<div class="grid" markdown>

- **`Conversation`**
    - This tab shows the main page of the PR.
    Here, you can find the original description of the PR along with any comments left on the PR, both reviewer comments and any comments you leave in response.
    - You can also find any [resolved comments](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/commenting-on-a-pull-request#resolving-conversations) here, if you need to refer to them again.
- **`Commits`**
    - This tab shows a list of all commits added to the branch that is being reviewed.
    Your reviewer may use this page to explore changes in individual commits.

<!-- comment to force two grid columns -->

- **`Checks`**
    - This tab shows results from certain types of code checks that have been automatically set up by the Data Lab; you can ignore this tab.
- **`Files changed`**
    - This tab shows a line-by-line overview of content you have added and removed in your branch.
    Changes shown here should reflect the difference between files in your branch and the base branch your PR is targeting.
This should be the `main` branch of the `AlexsLemonade/OpenScPCA-analysis` repository.
    - Referring to this tab is a great way to quickly see all your changes, as well as respond to [reviewer suggestions or comments](#in-line-comments).

@@

</div>

## Types of reviewer comments

When reviewing your pull request, there are several types of comments your reviewer may leave for you to respond to.

!!! tip "Reading review comments"
    Unresolved file-level and in-line review comments will appear both in the `Conversation` and in the `Files Changed` tabs.

@@

Your reviewer can leave file-specific comments that provide high-level feedback about the overall contents of a given file.

For example, this file-level review comment requests that the code author comment and document a script:

<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-filecomment.png){width="600"}
</figure>

### In-line comments

Your reviewer can also leave in-line comments with feedback about specific lines of code:

@@

- Alternatively, if there are multiple suggestions to accept, you can click the `Add suggestion to batch` button on each one.
_The `Add suggestion to batch` option is only available via the `Files changed` tab!_
    - Now, both suggestions are `Pending in batch` to be committed as a group:
<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-pending-batch.png){width="600"}
</figure>

You can commit the batched suggestions at once with the `Commit suggestions` button that will appear at the top of the `Files changed` tab:

<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-commitbatch.png){width="600"}
</figure>

Read this GitHub documentation for more about [applying suggested changes while responding to review](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/incorporating-feedback-in-your-pull-request#applying-suggested-changes).

In some cases a reviewer will request you make changes without providing in-line suggestions.
When this happens, you will need to make the code changes locally.
If you committed suggestions directly on GitHub before working on additional changes, be sure to sync your local branch with changes made on GitHub before changing any local code.

### 2024-04-25 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md

- You'll gain more context about why the reviewer left certain comments by reading the whole review first.
    - Several comments may also be related to each other, and taking in the full review will help you identify similar concepts or areas the reviewer noted.
- It's also possible that your reviewer misunderstood, or differently interpreted, some changes you made.
    - Reading the full review will help you understand if there are areas that need clarification before you can dive into responding.

## Navigating the pull request page

On the top of the PR page, you will see four tabs:

<figure markdown="span">
    ![Pull request tabs.](../../img/respond-to-review-01.png){width="800"}
</figure>

@@

### File-level comments

Your reviewer can leave file-specific comments that provide high-level feedback about the overall contents of a given file.

For example, this file-level review comment requests that the code author comment and document a script:

<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-02.png){width="600"}
</figure>

@@

### In-line comments

Your reviewer can also leave in-line comments with feedback about specific lines of code:

<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-03.png){width="600"}
</figure>

@@

You can directly reply to these review comments on GitHub.
For example, you might ask your reviewer to clarify a particular comment, or ask for more guidance about how to implement the review.
Or, you might just want to leave your reviewer a helpful note of some kind!

The way to responding to an existing comment differs depending on which PR tab you are leaving the comment from:

- If you are in the **`Files changed`** tab, use the button `Add single comment` to reply (_do not use `Start a review`_):
    <figure markdown="span">
        ![Reply to a review comment from the `Files changed` tab.](../../img/respond-to-review-04.png){width="600"}
    </figure>

- If you are in the **`Conversation`** tab, use the button `Comment`.
    <figure markdown="span">
        ![Reply to a review comment from the `Conversation` tab.](../../img/respond-to-review-05.png){width="600"}
    </figure>

@@

Reviewer comments can also include _suggestions_ for code changes, that you can incorporate into your PR directly on GitHub.

For example, below are two suggestions left as comments on an R script:

- In the first comment, the reviewer suggested that the code author add a new line to load the `readr` library.
- In the second comment, the reviewer suggested that the code author remove a line that loads the `dplyr` library.

<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-06.png){width="600"}
</figure>

@@

Suggestion comments have two associated buttons boxed in the image above: `Commit suggestion` and `Add suggestion to batch`.

- You can directly commit a _single_ suggestion by clicking `Commit suggestion` button, and then clicking `Commit changes` in the dropdown box:
<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-07.png){width="600"}
</figure>

- Alternatively, if there are multiple suggestions to accept, you can click the `Add suggestion to batch` button on each one.
_The `Add suggestion to batch` option is only available via the `Files changed` tab!_
    - Now, both suggestions are `Pending in batch` to be committed as a group:
<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-08.png){width="600"}
</figure>

@@

You can commit the batched suggestions at once with the `Commit suggestions` button that will appear at the top of the `Files changed` tab:

<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-09.png){width="600"}
</figure>

@@

## Responding to review comments

In some cases a reviewer will request you make changes without providing in-line suggestions.
When this happens, you will need to make the code changes locally.
If you committed suggestions directly on GitHub before working on additional changes, be sure to sync your local branch with changes made on GitHub before changing any local code.

When making code changes in response to review, you need to make sure your local repository is staying in-sync with any changes made directly on GitHub.
Follow these steps when responding to review to avoid [merge conflicts](../creating-pull-requests/resolve-merge-conflicts.md):

1. Before making any code changes, _pull_ your branch from GitHub.
This will ensure you are up-to-date with any code changes made on GitHub, including committed review suggestions or updates from merging the upstream `main` branch into your pull request branch.
    <figure markdown="span">
        ![Pull in GitKraken.](../../img/respond-to-review-10.png){width="600"}
    </figure>

@@

1. Make changes, commit, and push as you normally would.
Any changes you push to the branch with an open PR will automatically be reflected in the PR.
    - When committing changes in response to review, the smaller the better!
    Review will go more smoothly if each commit addresses only one review comment.

## Re-requesting review

Once you feel that you have addressed your reviewer's comments, you can [re-request review by clicking the cycle icon next to their handle on the right side of the PR page](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/incorporating-feedback-in-your-pull-request#re-requesting-a-review).

<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-11.png){width="400"}
</figure>

### 2024-04-25 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md

Your reviewer can also leave in-line comments with feedback about specific lines of code:

<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-03.png){width="600"}
</figure>

You can directly reply to these review comments on GitHub.
For example, you might ask your reviewer to clarify a particular comment, or ask for more guidance about how to implement the review.
Or, you might just want to leave your reviewer a helpful note of some kind!

The way to respond to an existing comment differs depending on which PR tab you are leaving the comment from:

@@

For example, below are two suggestions left as comments on an R script:

- In the first comment, the reviewer suggested that the code author add a new line to load the `readr` library.
- In the second comment, the reviewer suggested that the code author remove a line that loads the `dplyr` library.

<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-06.png){width="600"}
</figure>

### Responding to suggestion comments 
Suggestion comments have two associated buttons boxed in the image above: `Commit suggestion` and `Add suggestion to batch`.

@@

You can commit the batched suggestions at once with the `Commit suggestions` button that will appear at the top of the `Files changed` tab:

<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-09.png){width="600"}
</figure>

Read this GitHub documentation for more about [applying suggested changes while responding to review](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/incorporating-feedback-in-your-pull-request#applying-suggested-changes).

## Making additional code changes in response to comments

In some cases a reviewer will request you make changes to the code without providing in-line suggestions.
When this happens, you will need to make the code changes locally.
If you committed suggestions directly on GitHub before working on additional changes, be sure to sync your local branch with changes made on GitHub before changing any local code.

### 2024-04-25 docs/contributing-to-analyses/pr-review-and-merge/respond-to-review.md

<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-06.png){width="600"}
</figure>

## Responding to review comments

### Responding to suggestion comments
Suggestion comments have two associated buttons boxed in the image above: `Commit suggestion` and `Add suggestion to batch`.

- You can directly commit a _single_ suggestion by clicking `Commit suggestion` button.
Type a commit message, and then click `Commit changes` in the dropdown box:
<figure markdown="span">
    ![Re-request review.](../../img/respond-to-review-07.png){width="600"}
</figure>

### 2024-05-01 analyses/ewings-cell-type/02-cellassign-tumor-cell-classification.Rmd
initiate cellassign notebook

---
title: "Explore tumor cell classification with marker genes and CellAssign - SCPCS000490"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  sample_id: SCPCS000490
  library_id: SCPCL000822
---

@@

This notebook looks at expression of marker genes for tumor cells in a Ewing sarcoma sample, `r {params$sample_id}`. 
Additionally, we see if we can use those marker genes to classify tumor and normal cells using both manual annotation and `CellAssign`.

The first section focuses on assigning tumor and normal cells by looking at marker gene expression alone:

- First we look at expression of each of the marker genes across all cells. 
- Then we plot the mean expression of all marker genes across all cells. 
- Next, we perform clustering and see if we can assign each cluster as tumor or normal cells based on mean expression of marker genes.

The next section uses `CellAssign` to assign tumor cells with the same set of marker genes:

- We use `CellAssign` with both the full list of marker genes and filtered list of marker genes. 
- We then compare the average gene expression of all marker genes between tumor and other cells.

@@

marker_genes <- file.path(module_base, "references", "tumor_marker_genes.tsv")

# Output files
tumor_markers_mtx_file <- file.path(module_base, "references", "cellassign_refs", "tumor-marker-cellassign.tsv")
filtered_markers_mtx_file <- file.path(module_base, "references", "cellassign_refs", "filtered-tumor-marker-cellassign.tsv")
cellassign_predictions_file <- file.path(module_base, "results", "cellassign-tumor-predictions.tsv")
filtered_predictions_file <- file.path(module_base, "results", "filtered-cellassign-tumor-predictions.tsv")
```

@@

# read in marker genes table 
marker_genes_df <- readr::read_tsv(marker_genes) |> 
  # account for genes being from multiple sources
  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
  dplyr::distinct()

marker_genes <- marker_genes_df |> 
  dplyr::filter(cell_type == "tumor") |> 
  dplyr::pull(ensembl_gene_id)
```

@@

### Explore marker gene expression

The first thing we do here is just create a faceted UMAP showing the expression of each marker gene for tumor cells.

```{r}
# get the gene expression counts for all marker genes
marker_gene_exp <- logcounts(sce[marker_genes, ]) |>
    as.matrix() |>
    t() |>
    as.data.frame() |>
    tibble::rownames_to_column("barcodes")

# pull out the UMAP coordinates and make a data frame to use for plotting
umap_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP"),
        starts_with("UMAP")
      ) |> 
  # add in marker gene expression to dataframe 
  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
  # combine all genes into a single column for easy faceting
  tidyr::pivot_longer(
    cols = starts_with("ENSG"),
    names_to = "ensembl_gene_id",
    values_to = "gene_expression"
  ) |> 
  # join with marker gene df to get gene symbols for plotting 
  dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression, clusters)
```

```{r}
# faceted umap showing a umap panel for each marker gene 
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
    geom_point(alpha = 0.1, size = 0.2) +
    facet_wrap(vars(gene_symbol)) +
    scale_color_viridis_c() +
    labs(
      color = "Log-normalized gene expression"
    ) +
    # remove axis numbers and background grid
    scale_x_continuous(labels = NULL, breaks = NULL) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    theme(
      aspect.ratio = 1,
      legend.position = "bottom",
      axis.title = element_text(size = 9, color = "black"),
      strip.text = element_text(size = 8),
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8)
    ) +
    guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5)) +
  theme_bw()
```

Looking at this plot, there is definitely some variation in the expression of marker genes. 
It looks like the highest gene expression is seen in the larger group of cells on the bottom right of the UMAP.

Below, we can get the mean gene expression of all marker genes and plot in a single UMAP.

```{r}
# calculate mean gene expression across all marker genes in list 
marker_mean_exp <- umap_df |> 
  dplyr::group_by(barcodes) |> 
  dplyr::mutate(mean_exp = mean(gene_expression)) |> 
  dplyr::select(barcodes, UMAP1, UMAP2, mean_exp, clusters) |> 
  dplyr::distinct()

# plot mean gene expression 
ggplot(marker_mean_exp, aes(x = UMAP1, y = UMAP2, color = mean_exp)) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_color_viridis_c() +
  theme_bw()
```

Similar to the individual plots, it looks like there is one group of cells on the bottom right that has the highest mean gene expression. 
We would anticipate that these are most likely to be the tumor cells.

### Using clusters and marker gene expression to annotate tumor cells

We anticipate that the tumor cells and normal cells would cluster separately. 
First, let's just plot the cluster assignments alone.

```{r}
ggplot(marker_mean_exp, aes(x = UMAP1, y = UMAP2, color = clusters)) +
  geom_point(size = 0.5, alpha = 0.5) +
  theme_bw()
```

It looks like our large group of cells on the bottom right is comprised of a few clusters while most other groups of cells are a single cluster. 
Let's see if there's any difference in mean expression of the marker genes across the clusters. 
Based on the gene expression I would expect most cells in the bottom right to be considered tumor and all other cells to be normal.

```{r}
# plot marker gene expression across all clusters 
ggplot(marker_mean_exp, aes(x = mean_exp, fill = clusters)) +
  geom_density() +
  facet_grid(rows = vars(clusters)) +
  theme_bw()
```

Looking at this there are definitely distinct groups of cells with higher mean expression of the marker genes than other cells. 
Before doing anything fancy, let's use this to manually assign tumor and normal cells. 
One approach is to assign based on the cluster, so if the cluster has a median expression > 0.75 then all cells in that cluster would get assigned as tumor cells. 
We could also pick a mean expression cutoff and use that to assign each cell individually, but there are some tails that overlap.

```{r}
# classify clusters as tumor/normal 
cluster_classification_df <- marker_mean_exp |> 
  dplyr::group_by(clusters) |> 
  # calculate the median mean gene expression across all cells in a cluster
  dplyr::summarise(median_cluster_exp = median(mean_exp)) |> 
  # use 0.75 as the cutoff for tumor/normal 
  dplyr::mutate(tumor_normal_classification = dplyr::if_else(median_cluster_exp > 0.75, "Tumor", "Normal"))

# get individual cell classifications
celltype_classification_df <- marker_mean_exp |> 
  dplyr::left_join(cluster_classification_df)

# umap showing tumor and normal cells 
ggplot(celltype_classification_df, aes(x = UMAP1, y = UMAP2, color = tumor_normal_classification)) +
  geom_point(size = 0.5, alpha = 0.5) +
  theme_bw()
```

### Run CellAssign with marker genes

Now let's see if we can use an automated marker gene-based method to identify tumor cells. 
To do this we will use `CellAssign` with a reference matrix created from the marker gene list.

@@

```{r}
binary_mtx <- marker_genes_df |>
  dplyr::select(cell_type, ensembl_gene_id) |> 
  unique() |> 
  tidyr::pivot_wider(
    id_cols = ensembl_gene_id,
    names_from = cell_type,
    values_from = cell_type,
    values_fn = length,
    values_fill = 0
  ) |>
  # add a column with no marker genes
  # cell assign will assign cells to "other" when no other cell types are appropriate
  dplyr::mutate(other = 0)

# export matrix 
readr::write_tsv(binary_mtx, tumor_markers_mtx_file)
```

@@

```{r, eval=FALSE}
cellassign_call <- glue::glue("
python scripts/run-cellassign.py \
  --anndata_file '{anndata_file}' \
  --output_predictions '{cellassign_predictions_file}' \
  --reference '{tumor_markers_mtx_file}'
")

@@

```{r}
# read in predictions
predictions <- readr::read_tsv(cellassign_predictions_file)

# get individual cell type assignments
# those with the max prediction 
celltype_assignments <- predictions |>
      tidyr::pivot_longer(
        !barcode,
        names_to = "celltype",
        values_to = "prediction"
      ) |>
      dplyr::group_by(barcode) |>
      dplyr::slice_max(prediction, n = 1) |>
      dplyr::ungroup()
# create a table of assignments 
table(celltype_assignments$celltype)
```

@@

@@

# filter binary mtx 
filtered_binary_mtx <- binary_mtx |> 
  dplyr::filter(ensembl_gene_id %in% filtered_markers)

readr::write_tsv(filtered_binary_mtx, filtered_markers_mtx_file)
```

```{r, eval=FALSE}
cellassign_call <- glue::glue("
python scripts/run-cellassign.py --anndata_file '{anndata_file}' --output_predictions '{filtered_predictions_file}' --reference '{filtered_markers_mtx_file}'
")

@@

```{r}
predictions <- readr::read_tsv(filtered_predictions_file)

celltype_assignments <- predictions |>
      tidyr::pivot_longer(
        !barcode,
        names_to = "celltype",
        values_to = "prediction"
      ) |>
      dplyr::group_by(barcode) |>
      dplyr::slice_max(prediction, n = 1) |>
      dplyr::ungroup()

table(celltype_assignments$celltype)
```

@@

@@

- We do see variation in marker gene expression, with a large group of cells expressing multiple marker genes at a higher level than all other cells. 
- This same group of cells has the highest mean gene expression of all marker genes. 
- We see a clear separation in marker gene expression between clusters. 
- Assigning tumor cells by cluster identifies a distinct group of tumor cells. 
- Using `CellAssign` with the same list of marker genes doesn't identify any tumor cells. 
- Using `CellAssign` with a restricted list of marker genes identfies tumor cells that have lower marker gene expression than other cells. 
This is opposite of the assignment we obtained using manual annotation.

>>>>

---
title: "Explore tumor cell classification with CellAssign"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  sample_id: SCPCS000490
  library_id: SCPCL000822
---

@@

This notebook looks at classifying tumor cells with `CellAssign` in a Ewing sarcoma sample, `r {params$sample_id}`.

- We use `CellAssign` with both a full list of marker genes and filtered list of marker genes. 
- We then compare the average gene expression of all marker genes between tumor and other cells. 
- Finally we use `CellAssign` with marker genes from [Visser et al.](https://doi.org/10.1158/2767-9764.CRC-23-0027) specified in the supplemental methods.

@@

# tumor marker genes
tumor_markers_file <- file.path(module_base, "references", "tumor-marker-genes.tsv")
# all markers 
all_markers_file <- file.path(module_base, "references", "all-marker-genes.tsv")

# Output reference matrix files
ref_dir <- file.path(module_base, "references", "cellassign_refs")

tumor_markers_mtx_file <- file.path(ref_dir, "tumor-marker-cellassign.tsv")
filtered_markers_mtx_file <- file.path(ref_dir, "filtered-tumor-marker-cellassign.tsv")
visser_markers_mtx_file <- file.path(ref_dir, "visser-all-marker-cellassign.tsv")

# output predictions from cellassign 
results_dir <- file.path(module_base, "results")
cellassign_predictions_file <- file.path(results_dir, "cellassign-tumor-predictions.tsv")
filtered_predictions_file <- file.path(results_dir, "filtered-cellassign-tumor-predictions.tsv")
visser_predictions_file <- file.path(results_dir, "visser-cellassign-predictions.tsv")

```

@@

# read in tumor marker genes table 
tumor_markers_df <- readr::read_tsv(tumor_markers_file) |> 
  # account for genes being from multiple sources
  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
  dplyr::distinct()

# read in all maker genes 
all_markers_df <- readr::read_tsv(all_markers_file) |> 
  # account for genes being from multiple sources
  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
  dplyr::distinct()
```

Define some helper functions for creating a reference matrix and obtaining cell type annotations from `CellAssign` predictions.

```{r}
# function for creating a binary matrix with cell types as columns and marker genes as rows 
build_binary_mtx <- function(marker_genes_df,
                             gene_col, 
                             cell_type_col) {

  binary_mtx <- marker_genes_df |>
  dplyr::select({{cell_type_col}}, {{gene_col}}) |> 
  unique() |> 
  tidyr::pivot_wider(
    id_cols = {{gene_col}},
    names_from = {{cell_type_col}},
    values_from = {{cell_type_col}},
    values_fn = length,
    values_fill = 0
  ) |>
  # add a column with no marker genes
  # cell assign will assign cells to "other" when no other cell types are appropriate
  dplyr::mutate(other = 0)

  return(binary_mtx)
}

# function to get assigned cell type based on max prediction returned by CellAssign 
get_celltype_assignment <- function(predictions){
  # get individual cell type assignments
  # those with the max prediction 
  celltype_assignments <- predictions |>
    tidyr::pivot_longer(
      !barcode,
      names_to = "celltype",
      values_to = "prediction"
    ) |>
    dplyr::group_by(barcode) |>
    dplyr::slice_max(prediction, n = 1) |>
    dplyr::ungroup()

  return(celltype_assignments)
}
```

@@

### Run CellAssign with only tumor marker genes

First let's see if we can use `CellAssign` to classify cells as tumor or normal using just the list of tumor marker genes.

@@

```{r}
tumor_ref_mtx <- build_binary_mtx(tumor_markers_df, 
                                  gene_col = "ensembl_gene_id",
                                  cell_type_col = "cell_type")

# export matrix 
readr::write_tsv(tumor_ref_mtx, tumor_markers_mtx_file)
```

@@

```{r, eval=FALSE}
cellassign_call <- glue::glue(
  "python scripts/run-cellassign.py",
  "--anndata_file '{anndata_file}'",
  "--output_predictions '{cellassign_predictions_file}'",
  "--reference '{tumor_markers_mtx_file}'"
)

@@

```{r}
# read in predictions file
tumor_predictions <- readr::read_tsv(cellassign_predictions_file)

# get cell type assignments 
tumor_celltype_assignments <- get_celltype_assignment(tumor_predictions)

# create a table of assignments 
table(tumor_celltype_assignments$celltype)
```

@@

### Run CellAssign with a filtered set of tumor marker genes

@@

# filter binary mtx 
filtered_ref_mtx <- tumor_ref_mtx |> 
  dplyr::filter(ensembl_gene_id %in% filtered_markers)

readr::write_tsv(filtered_ref_mtx, filtered_markers_mtx_file)
```

```{r, eval=FALSE}
cellassign_call <- glue::glue(
  "python scripts/run-cellassign.py",
  "--anndata_file '{anndata_file}'",
  "--output_predictions '{filtered_predictions_file}'",
  "--reference '{filtered_markers_mtx_file}'"
)

@@

```{r}
filtered_predictions <- readr::read_tsv(filtered_predictions_file)

filtered_celltype_assignments <- get_celltype_assignment(filtered_predictions)

table(filtered_celltype_assignments$celltype)
```

@@

### Run CellAssign with tumor markers and all marker genes from Visser et al.

```{r}
visser_ref_mtx <- build_binary_mtx(all_markers_df, 
                                   gene_col = "ensembl_gene_id",
                                   cell_type_col = "cell_type")

# export matrix 
readr::write_tsv(visser_ref_mtx, visser_markers_mtx_file)
```

```{r, eval=FALSE}
cellassign_call <- glue::glue(
  "python scripts/run-cellassign.py ",
  "--anndata_file '{anndata_file}' ",
  "--output_predictions '{visser_predictions_file}' ",
  "--reference '{visser_markers_mtx_file}'"
)

```{r}
visser_predictions <- readr::read_tsv(visser_predictions_file)

visser_celltype_assignments <- get_celltype_assignment(visser_predictions)

table(visser_celltype_assignments$celltype)
```

@@

- Using `CellAssign` with the same list of marker genes doesn't identify any tumor cells. 
- Using `CellAssign` with a restricted list of marker genes identfies tumor cells that have lower marker gene expression than other cells. 
This is opposite of the assignment we obtained using manual annotation.

```{r}

```
